---
title: "PSYC234_Lab_Week 12_Student"
author: "Emma Mills"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# library(gdata)
library(PerformanceAnalytics)
library(effects)
library(gridExtra)
```


# Multiple Regression with Categorical Predictors

Many factors may predict academic success in young people that are not part of individual differences - i.e. their ability for a subject.  In this session, we are going to predict final Portuguese test scores (G3) from seven factors:

 - sex 
 - their health
 - their school
 - their school attendance rate
 - travel time to school
 - whether they attend extra private tutor classes
 - how much they drink at the weekend
 
Use the `Data_Codebook` .docx file in the lab folder to find out initial information such as variable names, their type and their levels or numerical ranges.


## Import Data

### Portuguese Scores

You are working with the "student-por.csv"


## Tidying

You could choose to reduce the dataset to keep only the variables they need; you don't need to do this - it is just about manageability.

If you want to do it, use the tidyverse `select()` function

Remember to also keep demographic variables so that you could describe your sample in your descriptive statistics results section

```{r}

```

## Visualise to get to know your data:

### Predictors

There is code in the lecture .Rmd file to help you do this for the categorical and continuous predictors.

Prompt questions to interrogate what you see:

 - Are the categorical variables balanced?
 - Are they made of two levels or more?
 - Do any of the categorical variables look like they might be ordinal?
 - How do the x-axes of the continuous variables look - does the range seem sensible?
 - Is there any skew to the continuous variables.
 
### Outcome variable

  - Is it a continuous variable?
  - Or if not, can we treat it as such for our test?

```{r message=FALSE, warning=FALSE}

```

Write yourself some text here as descriptors or what you see that helps you to get to know this unfamiliar dataset.


### Correlations betweeen continuous predictors

In the code below,  there is a bracket c( ).  You need to type the number of the columns that represent the continuous columns for which you want to see correlations - so what number column is the outcome predictor etc.  Separate each number with a comma.

```{r warning=FALSE}

chart.Correlation(d[, c( )], histogram=TRUE, pch=19)
```

Prompt questions to interrogate what you see:

thresholds to describe the relationships: 0.1 - 0.3 is small; 0.3 - 0.5 is medium; 0.5 - 0.7 is large.

  - What size of correlation do you see between the variables?
  - Do the correlations between predictors and the outcome variable suggest what you might see in the regression model?
  
Remember that these correlations are *bivariate* correlations - explaining shared variance between only those two predictors.  When we put the predictors into a regression model, we are estimating the independent variance of a predictor AFTER we have explained some of the variance already - in the other predictors. 

## Model

### Raw Variables

Write a model and call a summary of the model

```{r}

```
## Transforming Categorical Variables

Given the predictors, and having watched the lecture, what is the best decision you can make about how the categorical variables could be coded in the model?

Consider:

 - dummy coding scheme
 - sum coding scheme
 - transforming the catgorical variable into a numerical variable and centring
 - helmert coding
 
Make your decisions according to what is suggested by the model output and perform the transformations.  

If you want to use your script for revision,  it is worthwhile noting on the script why you are making the decisions that you are making - that way, if you submit for feedback, I can check your reasoning.

## Transforming Continuous Variables to Work with Categorical Variables

Consider:

 - raw variables
 - centred variables
 - standardised variables
 
In the lecture, I demonstrated how I label my variables according to the type of transformations that I make.  Make your decisions according to what is suggested by the model output and perform the transformations.  

I am not going to generate any more chunks in this script.  You need to do the rest.  

 - The keyboard shortcut to do that is Alt-Shift-I
 - The menu item is Code >> Insert Chunk
 
I strongly recommend that

 - you keep chunks relatively short
 - you run chunks after you have written code so that any error are identified in place and quickly - helping you to debug as you go along.

```{r}
```

## Model 

## Communicate

### Descriptive Statistics

You should report how many females / males,  the mean age of both sexes and how the sample was split by school.

### Report the Model

You should report:

Your model summary in a table - remember to refer to the Table in your textual description.

  - the name of the outcome variable
  - the names of the predictors in the model
  - then the omnibus test for the model (the F ratio, dfs and associated p value plus the amount of explained variance for hte outcome variable)
  - If you have a table, then you can report  the results of each significant predictor - i.e. report the levels of `walc` together, while referring the reader to the table for the other, non-significant results
  - If you choose not to make a table, then all the model results need to be reported in text
  - when you are reporting the statistical results of each predictor, make sure you give a verbal description of the result when the coefficient is significant.
  
### Plotting Predictions

You can communicate through plots also.  Plot some predictor effects that you could use in a results section. 

Use the screenshots of predictor effects as a self-check

Remember to refer to them as Figures in your textual descriptions.



